---
title: "sta 421 project (Life expectancy data analysis)"
author: "Silas Kibet"
date: '2022-11-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
GROUP 5


PROJECT OUTCOMES FOR THE ANALYSIS OF DATA ON DETERMINATION OF LIFE
EXPECTANCY

```{r}
expectancy<-read.csv(file.choose())
str(expectancy)
summary(expectancy)
```

From the summary statistics the data distributiom for this data set's
variables contains numerictype, one variable(status) which needs to be
converted to a dummy variable and missing values. Data cleaning

```{r}
x<-expectancy
sum(is.na(x))
colSums(is.na(x))
```

The number of missing values in the data set is 2563, which are contained
by 14 variables in the data set. This makes about 3% of all the data
points in this data set. The best way of dealing with these missing
values is by imputation. The missing values in the dataset will be
imputed using their respective median values.

```{r}
x[is.na(x$Hepatitis.B),"Hepatitis.B"]<-median(x$Hepatitis.B,na.rm=TRUE)
x[is.na(x$Polio),"Polio"]<-median(x$Polio,na.rm=TRUE)
x[is.na(x$Income.composition.of.resources),"Income.composition.of.resources"]<-median(x$Income.composition.of.resources,na.rm=TRUE)
x[is.na(x$BMI),"BMI"]<-median(x$BMI,na.rm=TRUE)
x[is.na(x$GDP),"GDP"]<-median(x$GDP,na.rm=TRUE)
x[is.na(x$Schooling),"Schooling"]<-median(x$Schooling,na.rm=TRUE)
x[is.na(x$polio),"Polio"]<-median(x$Polio,na.rm = TRUE)
x[is.na(x$Life.expectancy),"Life.expectancy"]<-median(x$Life.expectancy,na.rm=TRUE)
x[is.na(x$Total.expenditure),"Total.expenditure"]<-median(x$Total.expenditure,na.rm=TRUE)
x[is.na(x$thinness..1.19.years),"thinness..1.19.years"]<-median(x$thinness..1.19.years,na.rm=TRUE)
x[is.na(x$Adult.Mortality),"Adult.Mortality"]<-median(x$Adult.Mortality,na.rm=TRUE)
x[is.na(x$Diphtheria),"Diphtheria"]<-median(x$Diphtheria ,na.rm=TRUE)
x[is.na(x$thinness.5.9.years),"thinness.5.9.years"]<-median(x$thinness.5.9.years,na.rm=TRUE)
x[is.na(x$Population),"Population"]<-median(x$Population,na.rm=TRUE)
x[is.na(x$Alcohol),"Alcohol"]<-median(x$Population,na.rm=TRUE)
sum(is.na(x))
```

From the above output the sum of NA's in our dataset is now zero The
next thing is converting the status variable to a dummy variable

```{r}
x$Status<-ifelse(x$Status=="Developed",1,0)
head(x$Status)
```

The variable status is now in zeros and ones The next procedure inline
is carrying out data analysis 1.Identifying variables which are
significant

```{r}
lm(Life.expectancy~.,data=x)

```

Variables that are significant are: Year
Status
Adult.Mortality
infant.deaths
Alcohol
percentage.expenditure
Hepatitis.B Country 
#2. Should a country having a lower life expectancy
value(<65) increase its healthcare expenditure in order to improve its
average lifespan?

```{r}
library(ggplot2)
z<-x$Total.expenditure
y<-x$Life.expectancy
k<-subset(x,Life.expectancy<65,select=c(Life.expectancy,Total.expenditure))
ggplot(k,aes(y=Life.expectancy,x=Total.expenditure)) + geom_point(col=c("blue")) + ggtitle("A GRAPH OF TOTAL EXPENDITURE AGAINST LIFE EXPECTANCY") +xlab("EXPENDITURE") + ylab("LIFE EXPECTANCY") + geom_smooth(method="lm",col="black")
```

From the above plot an increase in expenditure negatively affects the
life expectancy. 3. How does Infant and Adult mortality rates affect
life expectancy.

```{r}
lm(Life.expectancy~infant.deaths+Adult.Mortality,data=x)#It negatively affects life expectancy.
```

```{r}
ggplot(x,aes(y=Life.expectancy,x=infant.deaths)) + geom_point(col=c("red")) +ggtitle("A GRAPH OF INFANT DEATHS AGAINST LIFE EXPECTANCY")+ xlab("INFANT DEATHS") + ylab("LIFE EXOECTANCY") + geom_smooth(method="lm",col="black")
```

Infant deaths negatively affects life expectancy as evident on the above
plot having a negative slope.

```{r}
lm(Life.expectancy~Adult.Mortality,data=x)
```

```{r}
ggplot(x,aes(y=Life.expectancy,x=Adult.Mortality)) + geom_point(col=c("purple")) +ggtitle("A PLOT OF ADULT MORTALITY AGAINST LIFE EXPECTANCY ") + xlab("ADULT MORTALITY") + ylab("LIFE EXPECTANCY") + geom_smooth(method="lm",col="black")

```

From the above plot and linear model. It is evident that Adult Mortality
reduces Life expectancy of individuals as a result of negative slope of
the graph and the negative intercept of Adult mortality. #4. Does Life
Expectancy has positive or negative correlation with eating habits,
lifestyle, exercise, smoking, drinking alcohol etc. used pearson
correlation to determine the correlation #correlation between Life
expectancy and BMI(for eating habits)

```{r}
cor.test(x$BMI ,x$Life.expectancy,alternative = c("two.sided","less","greater"),method="pearson",exact=NULL,conf.level=0.95,ccontinuity=FALSE)

```

For BMI and Life expectancy it exhibits a positive correlation though
weakly of 0.5569 #Correlation between life expectancy and alcohol

```{r}
cor.test(x$Alcohol,x$Life.expectancy,x$BMI ,alternative = c("greater"),method="pearson",exact=NULL,conf.level=0.95,ccontinuity=FALSE)#Variables are positively correlated
```

```{r}
cor.test(x$Alcohol,x$Life.expectancy,alternative = c("two.sided","less","greater"),method="pearson",exact=NULL,conf.level=0.95,ccontinuity=FALSE)

```

For Alcohol and Life expectancy there exists a positive correlation but
very weak, correlation of 0.022 #5. What is the impact of schooling on
the lifespan of humans?

```{r}
lm(Life.expectancy~Schooling,data=x)#It positively affects the lifespan of people/increases lifespan

```

Schooling increases the life expectancy of individuals by 2.077

```{r}
ggplot(x,aes(y=Life.expectancy,x=Schooling)) + geom_point(col=c("red")) +ggtitle("A GRAPH OF SCHOOLING AGAINST LIFE EXPECTANCY") + xlab("SCHOOLING") + ylab("LIFE EXPECTANCY") + geom_smooth(method="lm",col="black")
```

As evidentfrom the above plot with a positive slope, schooling increases
the life expectancy of indivduals. #6. Does Life Expectancy have
positive or negative relationship with drinking alcohol?

```{r}
cor.test(x$Alcohol,x$Life.expectancy,alternative = c("two.sided","less","greater"),method="pearson",exact=NULL,conf.level=0.95,ccontinuity=FALSE)

```

Alcohol and Life expectancy are positively correlated but weakly. #7. Do
densely populated countries tend to have lower life expectancy? #we must
scale the population variable first/normalization

```{r}
x$Population<-log(x$Population)
ggplot(x,aes(y=Life.expectancy,x=Population)) + geom_point(col=c("orange")) + geom_smooth(method="lm",col="red") + xlab("POPULATION") + ylab("LIFE EXPECTANCY") + ggtitle("A GRAPH OF POPULATION AGAINST LIFE EXPECTANCY")
```

From the above plot it is so clear that countries with low populations
have higher life expectancies as compared to over populated countries.
Though most of the people lie in the brackets of medium population. #8.
What is the impact of Immunization coverage on life Expectancy?

```{r}
summary(lm(Life.expectancy~Hepatitis.B+Polio+Measles+Diphtheria,data=x))

```

#BUILDING A PREDICTIVE MODEL TO EXPLAIN THE EFFECTS OF DRINKING ALCOHOL,HIV/AIDS,SCHOOLING,BMI,GDP,POLIO,MEASLES,HEPATITIS BSTATUS, ADULT MORTALITY INFANT DEATHS AND POPULATION ON LIFE EXPECTANCY.

```{r}
k<-subset(x,select=c(Life.expectancy,Alcohol,Status,Measles,Hepatitis.B,BMI,Polio,infant.deaths,HIV.AIDS,Schooling,GDP,Population,Adult.Mortality))

```
```{r}
library(corrplot)
correlation<- cor(x[,c(4,5,7,8,11,14,16,17,18,22)])
corrplot(correlation,method = "square")
```
#As per the corrplot we identify Adult Mortality,BMI,Schooling,HIV/AIDS to have great effects on the life expectancy.

#Loading required packages.
```{r}
library(tidyverse)
library(caret)
library(dplyr)
library(caTools)
```
#Splitting dataset into testing and training sets.
```{r}
set.seed(123)
training.sampl=sample.split(x$Life.expectancy,SplitRatio = 0.8)
train_data=subset(k,training.sampl==TRUE)
test_data=subset(k,training.sampl==FALSE)

```

#Building the model

```{r}
model1<-lm(Life.expectancy~.,data=train_data)
model1$coef
```

#making predictions

```{r}
predictions<-model1 %>% predict(test_data)
```

#Assessing model performance 
#using prediction error, RMSE

```{r}
#RMSE
RMSE(predictions,test_data$Life.expectancy)
#Mean
mean(test_data$Life.expectancy)
```
The prediction error RMSE is 4.262262 representing an error rate of 6.2% which is relatively low. Hence this is a good predictive model.

#Using R-square

```{r}
R2(predictions,test_data$Life.expectancy)
```

R2 value is 0.80 which implies 80 % of life expectancy can be explained by the predictor variables . This model is not bad in prediction.
